Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-07-09 Delaneyetal2024.tex
Overleaf work

\section{Results}
We first present formulations of sediment transport capacity in several channel types to build upon the analysis in \citet{alley1997}.
We then apply the models above (Section~\ref{sect:meth}) in two experiments by examining three model outputs.
Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).

@@

The first numerical experiment aims to compare the variability between the subglacial and subaerial model outputs to a range of channel shapes, friction factors, and water discharge variability, as the subglacial channel's cross-sectional area is allowed to evolve.
Following the findings in Section~\ref{sect:scaling}, variability in subglacial sediment discharge capacity could approach that of subaerial systems, should water discharge variations be sufficiently small.
For simplicity, we only examine $w\tau$.

@@

The greater variations in shear stress in subglacial channels presented here may actually cause even greater variations in sediment transport capacity in subglacial channels than presented above  (Figure~\ref{fig:multi_run}).
In sediment transport capacity relationships, such as in \citet{meyer1948} or \citet{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively (Section~\ref{sect:scaling}).
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).

@@

While transport-limited states at glaciers are likely rare \citep[e.g.][]{alley1997}, abundant sediment can persist underneath glaciers \citep[e.g.][]{walter2014,stevens2022,delaney2022}.
As a result, in some cases sediment export can be hydrologically controlled and efficient \citep[e.g.][]{swift2005,andresen2024}.
At these glaciers, the great variability in subglacial sediment transport capacity may make it difficult to link sediment discharge to hydrology, especially when peak events occur \citep{cowan1988,delaney2018,lu2022}.
As a result, the sediment transport capacity response to individual extreme precipitation or melt events could be difficult to establish.
The results here suggest that different channel sizes with the same hydrology forcing can result in very different sediment transport capacities.

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge.
High water discharge variability in \alpine{} may cause shear stress to approach $Q^{2}$in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} and stronger correlation between width-integrated shear stress may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

\competinginterests{The authors declare no competing interests.} %% this section is mandatory even if you declare that no competing interests are present

@@

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \citep{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
The evolving subglacial channel geometry presented in Section \ref{sect:sub_mode} is not considered.
Here we show the sediment transport capacity scaling with a given water discharge and hydraulic gradient for three different transport formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

@@

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results are summarised in Table~\ref{tab:eqs1}.
Now width integrated shear stress and sediment transport can be calculated for all regimes and all transport formulas.
The results of the 12 combinations are presented in Table~\ref{tab:Qs} and also in Table~S1 where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

\begin{table}[hbt!]
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel, and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \small
  \label{tab:eqs1}
  \begin{tabular}{llllll}
    Regime & Fixed & Determined via D-W
    & Additional relations & Shear stress & Stream p.\\
           & & (Equation~\ref{eq:DW})  &  &  \(\tau \propto\) & \(\Omega \propto\)\\
    \hline
    Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \makecell{\(w\,\propto Q^\alpha\) \\ \(S=wH\) \\ \(D_h\propto H\)} & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q\, \Psi\)\\
    R-channel & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q\, \Psi\)\\
    Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(fx Q^2 S^{-2}\) & \(f\, Q^3 S^{-5/2}\)\\
  \end{tabular}
\end{table}

@@

The sediment discharge in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.
However, the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely much higher in a steady-state R-channel.

Furthermore, R-channels will rarely operate in a steady state as variations in water discharge.
The timescale of diurnal melt or severe rain or melt is too short for the channel to reach a steady state.
In these cases with high water discharge variability, channels operate more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $1.7$ (Table~\ref{tab:Qs}).
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep{alley1997}.

>>>>

\section{Results}
%We then apply the models above (Section~\ref{sect:meth}) in two experiments by examining three model outputs.
%Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).

@@

The first numerical experiment aims to compare the variability between the subglacial and subaerial model outputs to a range of channel shapes, friction factors, and water discharge variability, as the subglacial channel's cross-sectional area is allowed to evolve.
%Following the findings in Section~\ref{sect:scaling}, variability in subglacial sediment discharge capacity could approach that of subaerial systems, should water discharge variations be sufficiently small.

@@

The greater variations in shear stress in subglacial channels presented here may actually cause even greater variations in sediment transport capacity in subglacial channels than presented above  (Figure~\ref{fig:multi_run}).
In sediment transport capacity relationships, such as in \citet{meyer1948} or \citet{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively (e.g. Section~\ref{sect:scaling}).
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).

@@

While transport-limited states at glaciers are likely rare \citep[e.g.][]{alley1997}, abundant sediment could persist underneath some glaciers \citep[e.g.][]{walter2014,stevens2022,delaney2022}.
As a result, sediment export can be hydrologically controlled and efficient \citep[e.g.][]{swift2005,andresen2024}.
At these glaciers, the great variability in subglacial sediment transport capacity may make it difficult to link sediment discharge to hydrology, especially when peak events occur \citep{cowan1988,delaney2018,lu2022}.
As a result, the sediment transport capacity response to individual extreme precipitation or melt events could be difficult to establish.
The results here suggest that different channel sizes with the same hydrology forcing can result in very different sediment transport capacities.

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge.
High water discharge variability in \alpine{} may cause shear stress to approach $Q^{2}$ in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} and stronger correlation between width-integrated shear stress may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

\competinginterests{The authors declare no competing interests.} %% This section is mandatory even if you declare that no competing interests are present

@@

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \citep{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Here we show the sediment transport capacity scaling with a given water discharge and hydraulic gradient for three different sediment transport capacity formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

@@

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results of are summarised in Table~\ref{tab:eqs1}.
Now width integrated shear stress and sediment transport can be calculated for all regimes and all transport formulas.
The results of the 12 combinations are presented in Table~\ref{tab:Qs} and also in Table~S1 where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

\begin{table}[hbt!]
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel, and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \small
  \label{tab:eqs1}
  \begin{tabular}{llllll}
    Regime & Fixed & Determined via D-W
    & Additional relations & Shear stress & Stream p.\\
           & & (Equation~\ref{eq:DW})  &  &  \(\tau \propto\) & \(\Omega \propto\)\\
    \hline
    Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \makecell{\(w\,\propto Q^\alpha\) \\ \(S=wH\) \\ \(D_h\propto H\)} & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q\, \Psi\)\\
    R-channel & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q\, \Psi\)\\
    Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f\, Q^3 S^{-5/2}\)\\
  \end{tabular}
\end{table}

@@

The sediment discharge capacity in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.
However, the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely much higher in a steady-state R-channel.

%Furthermore, R-channels will rarely operate in a steady state as variations in water discharge.
%The timescale of diurnal melt or severe rain or melt is too short for the channel to reach a steady state.
%In these cases with high water discharge variability, channels operate more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
%Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $1.7$ (Table~\ref{tab:Qs}).
%Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.
%These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep{alley1997}.
